Evaluation of Cache-Oblivious Data Structures\\Draft research proposal

### 2008-03-10 proposal.tex

\subsection{External memory model}
In the external memory model, as described by Aggarwal and Vitter (\cite{aggarwal1988ioc}), a distinction is made between internal memory (which is limited in size) and external memory (which is virtually unlimited). The external memory is subdivided into blocks of a fixed size and only entire blocks of data can be transferred between the two memories. The advantage of using such a model is that it more accurately reflects reality yet still allows formal analysis of the performance of various algorithms and data structures. Although Aggarwal and Vitter focus on magnetic disk as a storage medium for external memory, the model can be generalized to apply to every pair of adjacent levels in the memory hierarchy. In that case, the possibility of transferring several data blocks at once is usually dropped.

\subsection{Cache-oblivious memory model}
The external memory model can be simplified and extended to a cache-oblivious model (\cite{prokop1999coa}) which recognizes the same distinction between internal and external memory, but does not assume the relevant parameters (size of the cache and size of the data blocks) are known. Therefore, cache-oblivious data structures and algorithms must be defined without explicitly referencing these parameters, yet still perform well when evaluated in the cache-oblivious model. Algorithms that perform well in the cache-oblivious model are more general (since they have no parameters that must be specified) and by design tuned to all levels in the memory hierarchy, which means they should perform well in practice without any tuning.

@@

\subsection{Categories}
The development of the different memory models means data structures and algorithms can be classified depending on the assumptions about their environment used in their definitions.
\begin{list}{}{}
\item \textbf{Cache-aware} data structures and algorithms perform optimally in the external memory model, but require parametrization with some or all properties of the cache lay-out (such as block size or cache memory size).
\item \textbf{Cache-unaware} [does a better term exist?] data structures and algorithms  are designed with the uniform memory model in mind and may therefore not perform optimally in the external memory model.
\item \textbf{Cache-oblivious} data structures and algorithms are designed to perform optimally in the cache-oblivious model, meaning they are also optimal in the external memory model (in the asymptotic sense and up to a constant factor of memory transfers) without requiring tuning of parameters.
\end{list}

\section{Related work}
Many data structures and algorithms have been analyzed in the cache-oblivious model. In many cases, new data structures and algorithms have been introduced that perform optimally in this model. Prokop presents asymptotically optimal cache-oblivious algorithms for matrix transposition, fast Fourier transformation and sorting (\cite{prokop1999coa}). Bender, Demaine and Farach-Colton designed a cache-oblivious variant of a B-tree (\cite{bender2005cob}, \cite{bender2004lpc}).

@@

\section{Problem statement}
Several cache-oblivious data structures and algorithms have been proposed. In most cases, theoretical complexity analysis shows that the proposed solutions are asymptotically optimal, or equivalent to alternative non-oblivious solutions (up to a constant factor). However, in software engineering practice we are not only interesting in asymptotic computational complexity, but also in the practical performance of data structures and algorithms. Indeed, many algorithms that have undesirable theoretical properties are actually widely used because they perform well in practice (for example, sorting algorithms like Quicksort and Shell sort).

@@

From the application programmer's point of view, data structures and algorithms can be grouped together based on the functionality they provide. For example, all comparison-based sorting algorithms can be considered to provide equivalent functionality, even if their performance is different. Since it is not sensible to compare data structures which provide different functionality, it is necessary to limit the scope of the research to data structures that provide similar functions. For the proposed research, we will focus on dictionary-like data structures. Such data structures support storage and retrieval of data values by associating them with a particular data value, called a key.

@@

\subsection{Research questions}
Following from the above, the main question to be answered is:
\begin{list}{}{}
\item Do cache-oblivious data structures have practical merit?
\end{list}

@@

\section{Research approach}

@@

The goal of the research is to evaluate a variety of dictionary-like data structures in a practical setting. For this, we need a realistic scenario that makes heavy use of dictionaries, and that allows us to easily change the implementation used. The scenario that will be used, is that of a state-space search. The purpose of such an algorithm is to find an optimal path from a (set of) given initial state(s) to a goal state. State search algorithms are often used to solve reachability problems, which occur for example in formal model checking. The main data structures used are a queue (either a regular queue or, if search heuristics are used, a priority queue) and a dynamic set. An example of a simple state search algorithm is given in program \ref{prog-search}.

@@

Since searching elements in and adding elements to a dynamic set is more complicated than simply inserting and extracting states from a first-in-first-out queue, the execution time of the algorithm will be dominated by the performance of the set (assuming context-dependent functions like \verb#successors# and \verb#isGoal# in the sample code are efficient). The main focus of the research will therefore be on alternative implementations of the set data structure.

The situation changes, of course, when instead of a first-in-first-out queue a priority queue is used. This is often the case when we are searching for a shortest path (instead of any path that reaches the goal) or when heuristics are used in the search. In this case, the complexity of the queue increases. If time allows it, the research could be extended to evaluating different implementations of priority queues as well.

\subsection{Set operations}
A minimal set data structure should be able to determine whether or not a given element is a member of the set. A dynamic set differs from a static set in that it support operations to add and/or remove elements from the set dynamically. In our experiments, we will start out with an empty set and only allow for insertion of new elements. This means the our dynamic set must support two basic operations:
\begin{itemize}
\item \verb#insert(S,X)# -- Inserts $X$ into set $S$, if it is not yet present.
\item \verb#contains(S,X)# -- Returns whether set $S$ contains $X$.
\end{itemize}
In practice, it useful to be able to test if a set contains an element, and add this element if it does not. This can be implemented by first testing for the existence, and then conditionally inserting the element (as is done in program \ref{prog-search}) but many data structures allow to combine this into a single operation. We will therefore require that the \verb#insert# operation also returns whether or not the element was newly inserted.

\subsection{Set implementations}
Many different data structures have been developed that can be used to efficiently implement the operations described above. Several data structures will be selected from the three categories listed in the introduction:
\begin{itemize}
\item \textbf{Cache-unaware}: e.g. a hash table.
\item \textbf{Cache-aware}: e.g. a traditional B-tree.
\item \textbf{Cache-oblivious}: e.g. a cache-oblivious B-tree.
\end{itemize}
[To be determined: which data structures to use?]

@@

\subsection{Measurement}
As described in the problem statement, the main metric of interest is the total execution time of our algorithm, since the focus of the research is to determine the practical merit of the data structures used.

@@

\begin{table*}
\begin{center}
\begin{tabular}{ | l | l | l | }
\hline
\textbf{Task} & \textbf{Result} & \textbf{Deadline} \\
\hline
Finish research proposal
    & Final research proposal
    & March 27th \\
Study literature
    & "Related work" section
    & April 4th \\
Implement algorithms
    & Working code
    & April 28th \\
Finish paper contents
    & Complete paper for peer review
    & May 6th \\
Revise paper
    & Revised paper
    & May 23rd \\
Final paper formatting
    & Finished paper
    & June 11th \\
Prepare presentation
    & Presentation
    & June 20th \\
\hline
\end{tabular}
\caption{Planning}
\label{tab-planning}
\end{center}
\end{table*}

>>>>

\subsection{External Memory Model}
In the external memory model, as described by Aggarwal and Vitter (\cite{aggarwal1988ioc}), a distinction is made between internal memory (which is limited in size) and external memory (which is virtually unlimited). The external memory is subdivided into blocks of a fixed size and only entire blocks of data can be transferred between the two memories. The advantage of using such a model is that it more accurately reflects reality yet still allows formal analysis of the performance of various algorithms and data structures. Although Aggarwal and Vitter focus on magnetic disk as a storage medium for external memory, the model can be generalized to apply to every pair of adjacent levels in the memory hierarchy. In that case, the possibility of transferring several data blocks at once is usually dropped.

\subsection{Cache-Oblivious Memory Model}
The external memory model can be simplified and extended to a cache-oblivious model (\cite{prokop1999coa}) which recognizes the same distinction between internal and external memory, but does not assume the relevant parameters (size of the cache and size of the data blocks) are known. Therefore, cache-oblivious data structures and algorithms must be defined without explicitly referencing these parameters, yet still perform well when evaluated in the cache-oblivious model. Algorithms that perform well in the cache-oblivious model are more general (since they have no parameters that must be specified) and by design tuned to all levels in the memory hierarchy, which means they should perform well in practice without any tuning.

@@

\subsection{Categories}
The development of the different memory models means data structures and algorithms can be classified depending on the assumptions about their environment used in their definitions.
\begin{list}{}{}
\item \textbf{Cache-aware} data structures and algorithms perform optimally in the external memory model, but require parametrization with some or all properties of the cache layout (such as block size or cache memory size).
\item \textbf{Cache-unaware} [does a better term exist?] data structures and algorithms  are designed with the uniform memory model in mind and may therefore not perform optimally in the external memory model.
\item \textbf{Cache-oblivious} data structures and algorithms are designed to perform optimally in the cache-oblivious model, meaning they are also optimal in the external memory model (in the asymptotic sense and up to a constant factor of memory transfers) without requiring tuning of parameters.
\end{list}

\section{Related Work}
Many data structures and algorithms have been analyzed in the cache-oblivious model. In many cases, new data structures and algorithms have been introduced that perform optimally in this model. Prokop presents asymptotically optimal cache-oblivious algorithms for matrix transposition, fast Fourier transformation and sorting (\cite{prokop1999coa}). Bender, Demaine and Farach-Colton designed a cache-oblivious variant of a B-tree (\cite{bender2005cob}, \cite{bender2004lpc}).

@@

\section{Problem Statement}
Several cache-oblivious data structures and algorithms have been described above. Theoretical complexity analysis shows that the proposed solutions are asymptotically optimal. However, in software engineering practice we are not only interested in computational complexity, but also in the practical performance of data structures and algorithms. Indeed, many algorithms that have undesirable theoretical complexity are actually widely used because they perform well in practice (for example, sorting algorithms like Quicksort and Shell sort).

@@

From the application programmer's point of view, data structures and algorithms can be grouped together based on the functionality they provide. For example, all comparison-based sorting algorithms can be considered to provide equivalent functionality, even if their performance is different. Since it is not sensible to compare data structures which provide different functionality, it is necessary to limit the scope of the project to data structures that provide similar functions. For the proposed research, we will focus on dictionary-like data structures. Such data structures support storage and retrieval of data values by associating them with a particular data value, called a key.

@@

\subsection{Research Questions}
Following from the above, the main question to be answered is:
\begin{list}{}{}
\item Do cache-oblivious data structures have practical merit?
\end{list}

@@

\section{Research Approach}

@@

The goal of the project is to evaluate a variety of dictionary-like data structures in a practical setting. For this, we need a realistic scenario that makes heavy use of dictionaries, and that allows us to easily change the implementation used. The scenario that will be used, is that of a state-space search. The purpose of such an algorithm is to find an optimal path from a (set of) given initial state(s) to a goal state. State search algorithms are often used to solve reachability problems, which occur for example in formal model checking. The main data structures used are a queue (either a regular queue or, if search heuristics are used, a priority queue) and a dynamic set. An example of a simple state search algorithm is given in program \ref{prog-search}.

@@

Since searching elements in and adding elements to a dynamic set is more complicated than simply inserting and extracting states from a first-in-first-out queue, the execution time of the algorithm will be dominated by the performance of the set (assuming context-dependent functions like \verb#successors# and \verb#isGoal# in the sample code are efficient). The main focus of the research project will therefore be on alternative implementations of the set data structure.

The situation changes, of course, when instead of a first-in-first-out queue a priority queue is used. This is often the case when we are searching for a shortest path (instead of any path that reaches the goal) or when heuristics are used in the search. In this case, the complexity of the queue increases. If time allows it, the project could be extended to evaluating different implementations of priority queues as well.

\subsection{Set Operations}
A minimal set data structure should be able to determine whether or not a given element is a member of the set. A dynamic set differs from a static set in that it support operations to add and/or remove elements from the set dynamically. In our experiments, we will start out with an empty set and only allow for insertion of new elements. This means the our dynamic set must support two basic operations:
\begin{itemize}
\item \verb#insert(S,X)# -- Inserts $X$ into set $S$, if it is not yet present.
\item \verb#contains(S,X)# -- Returns whether set $S$ contains $X$.
\end{itemize}
In practice, it useful to be able to test if a set contains an element, and add this element if it does not. This can be implemented by first testing for the existence, and then conditionally inserting the element (as is done in program \ref{prog-search}) but many data structures allow to combine this into a single operation. We will therefore require that the \verb#insert# operation also returns whether or not the element was newly inserted.

\subsection{Set Implementations}
Many different data structures have been developed that can be used to efficiently implement the operations described above. Several data structures will be selected from the three categories listed in the introduction:
\begin{itemize}
\item \textbf{Cache-unaware}: e.g. a hash table.
\item \textbf{Cache-aware}: e.g. a traditional B-tree.
\item \textbf{Cache-oblivious}: e.g. a cache-oblivious B-tree.
\end{itemize}
[To be determined: which data structures to use?]

@@

\subsection{Measurement}
As described in the problem statement, the main metric of interest is the total execution time of our algorithm, since the focus of the project is to determine the practical merit of the data structures used.

@@

\begin{table*}
\begin{center}
\begin{tabular}{ | l | l | l | }
\hline
\textbf{Task} & \textbf{Result} & \textbf{Deadline} \\
\hline
Finish research proposal
    & Final research proposal
    & March 27th \\
Study literature
    & "Related Work" section
    & April 4th \\
Implement algorithms
    & Working code
    & April 28th \\
Finish paper contents
    & Complete paper for peer review
    & May 6th \\
Revise paper
    & Revised paper
    & May 23rd \\
Final paper formatting
    & Finished paper
    & June 11th \\
Prepare presentation
    & Presentation
    & June 20th \\
\hline
\end{tabular}
\caption{Planning}
\label{tab-planning}
\end{center}
\end{table*}
